Ecosystem functional response across precipitation extremes in a sagebrush steppe

### 2018-01-18 manuscript/Appendix4.Rmd
adds SOILWAT appendix and updates ordering in main text. Closes #13

\begin{center}
\textbf{\Large{Appendix 4}} \\
A.T. Tredennick, A.R. Kleinhesselink, B. Taylor \& P.B. Adler \\
``Consistent ecosystem functional response across precipitation extremes in a sagebrush steppe'' \\
\emph{PeerJ}
\end{center}

\section{Characterizing Extreme Precipitation Amounts}

Following the proposed methods of @Lemoine2016, we calculated quantiles from the empirical distribution of growing season precipitation at Dubios, ID.
We chose the 1\% quantile to be indicative of extreme dry conditions (drought) and the 99\% quantile to be indicative of extreme wet conditions (irrigation).
The data consist of 91 yearly records, which we assume are approximately normally distributed.
The R code below shows our procedure, and Fig. A3-1 shows the results.

```{r hrv, fig.cap="Density of the empirical distribution of growing season precipitation at Dubois, ID. Dashed vertical lines show the 1% and 99% quantiles, assuming a normal distribution."}
library(ggplot2)
weather    <- read.csv("../data/weather/ClimateIPM.csv")
mean_ppt   <- mean(weather$ppt1)
quants_ppt <- quantile(weather$ppt1,probs = c(0.01,0.99))
quants_ppt[1]/mean_ppt*100 # percent of mean ppt for drought
quants_ppt[2]/mean_ppt*100 # percent of mean ppt for irrigation

ggplot(weather, aes(x=ppt1))+
  geom_histogram(bins=20, color="dodgerblue", fill="dodgerblue", aes(y=..density..))+
  geom_line(stat="density", color="blue")+
  geom_vline(aes(xintercept=quants_ppt[1]), linetype=2)+
  geom_vline(aes(xintercept=quants_ppt[2]), linetype=2)+
  ylab("Density")+
  xlab("Growing Season Precipitation (mm)")+
  theme_bw()+
  theme(panel.grid.minor = element_blank())
```

>>>>

\begin{center}
\textbf{\Large{Appendix 4}} \\
A.T. Tredennick, A.R. Kleinhesselink, J.B. Taylor \& P.B. Adler \\
``Consistent ecosystem functional response across precipitation extremes in a sagebrush steppe'' \\
\emph{PeerJ}
\end{center}

\section{Random Slopes, Random Intercepts Model}
\subsection{Model Description}
Our hierarchical Bayesian model allows us to test for differences among treatments in the relationship between ANPP and soil moisture, and allowed us to account for the non-independence of observations through time within a plot.
Treatment differences are modeled as fixed effects, which are modified by plot-level random effects.
In what follows, $\textbf{X}$ is the fixed effects design matrix including:

1. a column of 1s for intercepts
2. a column of continuous values for the scaled volumetric water content for each observation
3. a column of 0s or 1s indicating whether the observation is from a drought treatment 
4. a column of 0s or 1s indicating whether the observation is from an irrigation treatment
5. a column of continuous values for the interaction between scaled volumetric water content and the drought treatment indicator (column 2)
6. a column of continuous values for the interaction between scaled volumetric water content and the irrigation treatment indicator (column 3)

The first seven rows of the fixed effects design matrix (six control plots and one drought plot) is as follows:

```{r fixed_design, echo=F, eval=T, message=F, warning=F}
library(tidyverse)    # Data munging
library(dplyr)        # Data summarizing
library(ggthemes)     # Pleasing ggplot themes
library(ggjoy)        # Joy plots!
library(stringr)      # Working with strings
library(rstan)        # For MCMC and Stan objects
library(gridExtra)    # For combining ggplot objects
library(viridis)      # Pleasing color palette
library(RColorBrewer) # More color palettes
library(cowplot)      # For combining ggplots
source("../code/read_format_data.R") # load data
lmod <- lm(log(anpp) ~ vwc_scaled*Treatment, anpp_data)
x    <- model.matrix(lmod) 
colnames(x) <- c("Int", "VWC", "Drought", "Irrig", "Drought:VWC", "Irrig:VWC")
head(x, 7)
```

\noindent{}Note that within a year, all plots within a treatment share the same value of volumetric water content. This is because we could not monitor soil moisture in each plot, and instead used sparse observations to model average soil moisture in each treatment in each year (see main text). Using this design matrix, we can estimate six fixed effects ($\boldsymbol{\beta}$s):

1. the intercept of the soil moisture-ANPP relationship for control plots
2. the slope of the soil moisture-ANPP relationship for control plots
3. the intercept offset for drought plots
4. the intercept offset for irrigation plots
5. the slope offset for drought plots
6. the slope offset for irrigation plots

\noindent{}We are particularly interested in the slope offsets, because these allow us to test whether the slopes for drought or irrigation are different from the control slope. If the slope offset is different from zero, this indicates the slopes are different. We assess whether the slope offsets are different from zero by calculating the probability that the posterior distribution is less or greater than zero (one-tailed tests). If the probability is greater than 0.95, then there is strong evidence that the slope offset is different from zero, and thus different from the control treatment slope.

To account for the fact that observations within plots through time are not independent, we include random effects that modify the fixed effects in each plot. We model these random effects ($\boldsymbol{\gamma}$s) as offsets drawn from a multivariate normal distribution with mean 0 and a variance-covariance matrix ($\Sigma$) that includes a covariance between the intercept and slope offsets. We implement this random effects structure by including a random effects design matrix ($\textbf{Z}$) with a column of 1s for intercept offsets and a column of continuous values for the volumetric water content for each observation.

Lastly, to account for unkown variation across years, we include random year effects. These year effects ($\boldsymbol{\eta}$s) act as offsets on the intercept.

Putting it all together, our model is defined mathematically as follows, where *i* denotes observation, *j* denotes plot, and *t* denotes year.
We assume the observations are conditionally Gaussian,

\vspace{-2em}

\begin{align}
\textbf{y} &\sim \text{Normal} \left(\boldsymbol{\mu}, \sigma^2 \right),
\end{align}

\noindent{}where $\boldsymbol{\mu}$ is the determinstic expectation from the regression model,

\vspace{-2em}

\begin{align}
\mu_{i} &= \boldsymbol{\beta}\textbf{x}_i + \boldsymbol{\gamma}_{j(i)}\textbf{z}_i + \eta_t.
\end{align}

\noindent{}All fixed effect $\beta$s were drawn from normally distributed priors with mean 0 and standard deviation 5: $\boldsymbol{\beta} \sim \text{Normal}(0,5)$. $\gamma$ random effects were drawn from a multivariate prior centered on zero with a shared variance covariance matrix:

\vspace{-2em}

\begin{align}
\boldsymbol{\gamma} \sim \text{MVN} \left( 0, \Sigma \right), \\
\Sigma = \sigma_{plt}^2 \textbf{R}, \\
\sigma_{plt} \sim \text{Cauchy}(0,2.5), \\
\textbf{R} \sim \text{LKJ}(2.0)
\end{align}

\noindent{}The random year effects ($\boldsymbol{\eta}$) are modeled as intercept offsets centered on zero with a shared variance ($\sigma_{yr}$): $\boldsymbol{\gamma} \sim \text{Normal} \left( 0, \sigma_{yr}  \right)$.

The Bayesian posterior distribution of our model can be expressed as:

\vspace{-2em}

\begin{align}
\left[ \boldsymbol{\beta}, \boldsymbol{\gamma}, \boldsymbol{\eta}, \sigma_{yr}, \sigma_{plt}, \textbf{R}, \sigma \right | \textbf{y}] &\propto \left( \prod_{i=1}^{n} \left[ y_i | \boldsymbol{\beta}, \boldsymbol{\gamma}, \boldsymbol{\eta}, \sigma \right] \right) \left( \prod_{j=1}^{J} \left[ \gamma_j | \sigma_{plt}, \textbf{R} \right] \right) \\
&\times \left( \prod_{t=1}^{T} \left[ \eta_t | \sigma_{yr} \right] \right)  [\boldsymbol{\beta}] [\sigma_{plt}] [\textbf{R}] [\sigma_{yr}] [\sigma]
\end{align}

We fit the model using MCMC as implemented in the software Stan [@stan2016].
Our Stan code is below.
All code necessary to reproduce our results has been archived on Figshare (*link here*) and released on GitHub (https://github.com/atredennick/usses_water/releases/v0.1).

\newpage{}

\subsection{Stan Code}
```{r stan1, echo=T, eval=F}
data {
  int<lower=0> Npreds;         # number of covariates, including intercept
  int<lower=0> Npreds2;        # number of random effect covariates
  int<lower=0> Nplots;         # number of plots
  int<lower=0> Ntreats;        # number of treatments
  int<lower=0> Nobs;           # number of observations
  int<lower=0> Nyears;         # number of years
  vector[Nobs] y;              # vector of observations
  row_vector[Npreds] x[Nobs];	 # design matrix for fixed effects
  row_vector[Npreds2] z[Nobs]; # simple design matrix for random effects 
  int plot_id[Nobs];           # vector of plot ids
  int treat_id[Nobs];          # vector of treatment ids
  int year_id[Nobs];           # vector of year ids
}

parameters {
	vector[Npreds] beta;	             # overall coefficients
	vector[Nyears] year_off;           # vector of year effects
	cholesky_factor_corr[Npreds2] L_u; # cholesky factor of plot random effect corr matrix
  vector[Npreds2] beta_plot[Nplots]; # plot level random effects
	vector<lower=0>[Npreds2] sigma_u;  # plot random effect std. dev.
	real<lower=0> sd_y;                # treatment-level observation std. dev.
	real<lower=0> sigma_year;          # year std. dev. hyperprior
}

transformed parameters {
  vector[Nobs] yhat;               # vector of expected values
  vector[Npreds2] u[Nplots];       # transformed plot random effects
  matrix[Npreds2,Npreds2] Sigma_u; # plot ranef cov matrix

  Sigma_u = diag_pre_multiply(sigma_u, L_u); # plot-level covariance matrix 
  for(j in 1:Nplots)
    u[j] = Sigma_u * beta_plot[j]; # plot random intercepts and slopes

  # regression model for expected values (one for each plot-year)
  for (i in 1:Nobs)
    yhat[i] = x[i]*beta + z[i]*u[plot_id[i]] + year_off[year_id[i]]; 
}

model {
  ####  PRIORS
  sigma_u ~ cauchy(0,2.5)
  sigma_year ~ cauchy(0,2.5)
  year_off ~ normal(0,sigma_year); # priors on year effects, shared variance
  beta ~ normal(0,5);	             # priors on treatment coefficients
  L_u ~ lkj_corr_cholesky(2.0);      # prior on the cholesky factor which controls the 
                                   # correlation between plot level treatment effects

  for(i in 1:Nplots)
		beta_plot[i] ~ normal(0,1); # plot-level coefficients for intercept and slope

	####  LIKELIHOOD
  for(i in 1:Nobs)
    y[i] ~ normal(yhat[i], sd_y); # observations vary normally around expected values
}

generated quantities{
  corr_matrix[Npreds2] R = multiply_lower_tri_self_transpose(L_u);
  cov_matrix[Npreds2] V = quad_form_diag(R,sigma_u);
}
```

\newpage{}

# Figures

```{r scatterplot, fig.cap="Scatterplot, on the arithmetic scale, of the data and model estimates shown as solid lines. Model estimates come from treatment level coefficients (colored lines).", echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)    # Data munging
library(dplyr)        # Data summarizing
library(ggthemes)     # Pleasing ggplot themes
library(ggjoy)        # Joy plots!
library(stringr)      # Working with strings
library(rstan)        # For MCMC and Stan objects
library(gridExtra)    # For combining ggplot objects
library(viridis)      # Pleasing color palette
library(RColorBrewer) # More color palettes
library(cowplot)      # For combining ggplots

source("../code/read_format_data.R") # load data
all_fit <- readRDS("../results/randcoefs_alltreatments_fit.RDS")

param_labels <- c("Control::Intercept",
                  "Drought::Intercept", 
                  "Irrigation::Intercept",
                  "Control::Slope", 
                  "Drought::Slope", 
                  "Irrigation::Slope")

treat_effects <- data.frame(extract(all_fit, pars = 'beta')) %>% 
  mutate(iteration = row_number()) %>% 
  gather(param_name, estimate, starts_with('beta')) %>%
  mutate(param_name = factor(param_name, labels = param_labels)) %>% 
  spread(param_name, estimate) %>% 
  mutate(`Drought::Intercept` = `Drought::Intercept` + `Control::Intercept`, 
         `Irrigation::Intercept` = `Irrigation::Intercept` + `Control::Intercept`,  
         `Drought::Slope` = `Drought::Slope` + `Control::Slope`, 
         `Irrigation::Slope` = `Irrigation::Slope` + `Control::Slope`) %>%
  gather(param_name, estimate, -iteration)

mean_effects <- treat_effects %>%
  group_by(param_name) %>%
  summarise(mean_estimate = mean(estimate))

regression <- mean_effects %>%
  separate(param_name, into = c("Treatment","Coefficient"), sep = "::") %>%
  spread(Coefficient, mean_estimate)

lmod <- lm(log(anpp) ~ vwc_scaled, anpp_data)
x    <- model.matrix(lmod) 
newx <- data.frame(x = seq(min(x[,2]), max(x[,2]), length.out = 50)) %>%
  mutate(id = paste("x",1:50,sep = "::")) %>%
  spread(id, x)
newx <- rbind(rbind(newx,newx),newx)
newx$Treatment <- regression$Treatment

regression <- regression %>%
  left_join(newx, by="Treatment") %>%
  gather(xid, newx, -Treatment, -Intercept, -Slope)
regression$estimate <- with(regression, Intercept+Slope*newx)

mean_log_anpp <- mean(log(anpp_data$anpp))
sd_log_anpp <- sd(log(anpp_data$anpp))
mean_vwc <- mean(anpp_data$total_seasonal_vwc)
sd_vwc <- sd(anpp_data$total_seasonal_vwc)

regression <- regression %>%
  mutate(backtrans_estimate = estimate*sd_log_anpp+mean_log_anpp,
         backtrans_vwc = newx*sd_vwc+mean_vwc)

regression_limited <- {}
for(dotrt in unique(regression$Treatment)){
  tmpdat <- filter(regression, Treatment == dotrt)
  tmprange <- range(anpp_data[which(anpp_data$Treatment==dotrt),"total_seasonal_vwc"])
  tmpdat$backtrans_vwc[tmpdat$backtrans_vwc < tmprange[1] | tmpdat$backtrans_vwc > tmprange[2]] <- NA
  regression_limited <- rbind(regression_limited, tmpdat)
}

ggplot(anpp_data, aes(x=total_seasonal_vwc,y=anpp))+
  geom_point(shape=21,color="grey25",alpha=0.8,aes(fill=Treatment))+
  geom_line(data=regression_limited, aes(x=backtrans_vwc, y=exp(backtrans_estimate), color=Treatment), size=1)+
  scale_fill_brewer(palette = "Set2")+
  scale_color_brewer(palette = "Set2")+
  xlab("March-June Cumulative VWC")+
  ylab(expression(paste("ANPP (g ",m^2,")")))+
  theme_bw()+
  theme(panel.grid.minor = element_blank())

```

\newpage{}

```{r posteriors, fig.cap="Prior (dashed lines) and posterior (solid line) distributions of intercepts and slopes for each treatment. The slope represents the main effect of soil moisture on log(ANPP). The red line marks 0. Shrinkage of the posterior distribution and/or changes in the mean indicate the data informed model parameters beyond the information contained in the prior for all coefficients.",echo=FALSE, message=FALSE, warning=FALSE}
posteriors <- treat_effects %>%
  separate(param_name, into = c("Treatment", "Coefficient")) %>%
  mutate(prior = rnorm(n = nrow(treat_effects), 0, 5)) %>%
  gather(Distribution, Value, -iteration, -Treatment, -Coefficient)

ggplot(posteriors, aes(x=Value, linetype=Distribution))+
    geom_line(stat = "density")+
    geom_vline(aes(xintercept=0), color="red", size=0.2)+
    scale_x_continuous(limits=c(-5,5))+
    scale_linetype_discrete(labels = c("Posterior","Prior"), name = NULL)+
    facet_grid(Coefficient~Treatment)+
    ylab("Density")+
    theme_few()+
    theme(axis.text = element_text(size=7))
```

\newpage{}

```{r year_effects, fig.cap="Posterior distributions of random year effects (intercept offsets). Kernel bandwidths of posterior densities were adjusted by a factor of 5 to smooth the density curves for visual clarity.", echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)    # Data munging
library(dplyr)        # Data summarizing
library(ggthemes)     # Pleasing ggplot themes
library(ggjoy)        # Joy plots!
library(stringr)      # Working with strings
library(rstan)        # For MCMC and Stan objects
library(gridExtra)    # For combining ggplot objects
library(viridis)      # Pleasing color palette
library(RColorBrewer) # More color palettes
library(cowplot)      # For combining ggplots

all_fit <- readRDS("../results/randcoefs_alltreatments_fit.RDS")

year_id_names <- data.frame(year_id = c(1:5),
                            year_name = c(1:5))
betas <- reshape2::melt(rstan::extract(all_fit, pars="year_off")) %>%
  rename(iteration = iterations, year_id = Var2, estimate = value, stan_name = L1) %>%
  left_join(year_id_names, by="year_id")

ggplot(betas, aes(x=estimate, y=as.factor(year_id), fill=as.factor(year_id), color=as.factor(year_id), height = ..density..))+
  geom_vline(aes(xintercept=0), linetype=2, color="grey45")+
  geom_joy(stat="density", adjust=3, alpha=0.9)+
  scale_x_continuous(breaks=seq(-3,3,0.5), limits=c(-3,3))+
  scale_y_discrete(labels = c("2012","2013","2014","2015","2016"))+
  scale_fill_viridis(end=0.8,discrete=T)+
  scale_color_viridis(end=0.8,discrete=T)+
  guides(fill=FALSE, color=FALSE)+
  xlab("Parameter Value")+
  ylab("Year")+
  theme_bw()+
  theme(panel.grid.minor = element_blank(),
        panel.border     = element_blank(),
        axis.line        = element_blank(),
        axis.ticks       = element_blank())

```

\newpage{}

### 2018-01-18 manuscript/Appendix5.Rmd
adds SOILWAT appendix and updates ordering in main text. Closes #13

\begin{center}
\textbf{\Large{Appendix 5}} \\
A.T. Tredennick, A.R. Kleinhesselink, J.B. Taylor \& P.B. Adler \\
``Consistent ecosystem functional response across precipitation extremes in a sagebrush steppe'' \\
\emph{PeerJ}
\end{center}

\section{Characterizing Extreme Precipitation Amounts}

Following the proposed methods of @Lemoine2016, we calculated quantiles from the empirical distribution of growing season precipitation at Dubios, ID.
We chose the 1\% quantile to be indicative of extreme dry conditions (drought) and the 99\% quantile to be indicative of extreme wet conditions (irrigation).
The data consist of 91 yearly records, which we assume are approximately normally distributed.
The R code below shows our procedure, and Fig. A5-1 shows the results.

```{r hrv, fig.cap="Density of the empirical distribution of growing season precipitation at Dubois, ID. Dashed vertical lines show the 1% and 99% quantiles, assuming a normal distribution."}
library(ggplot2)
weather    <- read.csv("../data/weather/ClimateIPM.csv")
mean_ppt   <- mean(weather$ppt1)
quants_ppt <- quantile(weather$ppt1,probs = c(0.01,0.99))
quants_ppt[1]/mean_ppt*100 # percent of mean ppt for drought
quants_ppt[2]/mean_ppt*100 # percent of mean ppt for irrigation

ggplot(weather, aes(x=ppt1))+
  geom_histogram(bins=20, color="dodgerblue", fill="dodgerblue", aes(y=..density..))+
  geom_line(stat="density", color="blue")+
  geom_vline(aes(xintercept=quants_ppt[1]), linetype=2)+
  geom_vline(aes(xintercept=quants_ppt[2]), linetype=2)+
  ylab("Density")+
  xlab("Growing Season Precipitation (mm)")+
  theme_bw()+
  theme(panel.grid.minor = element_blank())
```

# References

>>>>

\begin{center}
\textbf{\Large{Appendix 5}} \\
A.T. Tredennick, A.R. Kleinhesselink, B. Taylor \& P.B. Adler \\
``Consistent ecosystem functional response across precipitation extremes in a sagebrush steppe'' \\
\emph{PeerJ}
\end{center}

\section{Conducting our analysis with NDVI}

The first step of our analysis is to convert NDVI from field-based radiometric measurements to aboveground net primary productivity.
However, this conversion is not perfect and the associated uncertainty is sometimes high (Table A1-2).
Therefore, we also assessed ecosystem functional response across treatments using NDVI as the response variable rather than ANPP.
The model essentially the same as the ANPP model, except we use a Beta likelihood for NDVI since its values range from 0 to 1, not including true 0s or 1s (see Stan code, below).

Our NDVI-model results are very similar to the ANPP-model results (Figure A5-1).
As with our ANPP-model results, the positive offset for the drought treatment does not result in a significantly different slope once applied to the control slope.

```{r plot_betas, fig.cap="Posterior distributions for the effects of drought and irrigation on the intercept and slope of the NDVI-soil moisture relationship. Treatment effects show the difference between the coefficients estimated in the treated plots and the control plots. Probabilities (Pr </> 0 =) for each coefficient indicate the one-tailed probability that the coefficient is less than or greater than zero, depending on whether the median of the distribution is less than or greater than zero. The posterior densities were smoothed for visual clarity by increasing kernel bandwidth by a factor of five.", include=T, echo=F, comment=FALSE,warning=FALSE,message=FALSE,results='asis'}

library(tidyverse)    # Data munging
library(dplyr)        # Data summarizing
library(ggthemes)     # Pleasing ggplot themes
library(ggjoy)        # Joy plots!
library(stringr)      # Working with strings
library(rstan)        # For MCMC and Stan objects
library(gridExtra)    # For combining ggplot objects
library(viridis)      # Pleasing color palette
library(RColorBrewer) # More color palettes
library(cowplot)      # For combining ggplots

all_fit <- readRDS("../results/randcoefs_alltreatments_NDVIfit.RDS")

####
####  FUNCTION TO CALCULATE MAX(Pr(effect) > 0, Pr(effect) < 0) ----
####
get_one_tailed <- function(values){
  above <- 1 - ecdf(values)(0)
  below <- ecdf(values)(0)
  max(above,below)
}

####
####  PLOT TREATMENT-LEVEL POSTERIOR DISTRIBUTIONS ----
####
mycols <- c("#009E73", "#D55E00", "#0072B2")
param_labels <- c("Control::Intercept",
                  "Drought::Intercept", 
                  "Irrigation::Intercept",
                  "Control::Slope", 
                  "Drought::Slope", 
                  "Irrigation::Slope")

betas <- data.frame(extract(all_fit, pars = 'beta')) %>% 
  mutate(iteration = row_number()) %>% 
  gather(param_name, estimate, starts_with('beta')) %>%
  mutate(param_name = factor(param_name, labels = param_labels)) %>%
  separate(param_name, c("Treatment", "Type"), "::")

treat_slopes <- betas %>%
  filter(Treatment != "Control") %>%
  mutate(Type = paste(Type,"Offset"))

slope_probs <- treat_slopes %>%
  group_by(Treatment, Type) %>%
  summarise(probs = round(get_one_tailed(estimate),2)) %>%
  mutate(prob_text = paste("Pr < 0 =", probs))
slope_probs$prob_text[1] <- paste("Pr > 0 =", slope_probs$probs[1]) # switch direction of drought intercpet since greater than 0
slope_probs$xpos <- c(-2.8, -0.6, -2.6, -0.6)
slope_probs$ypos <- c(1.3,1.3,2.55,2.7)

treat_cols <- mycols[2:3]

ggplot(treat_slopes)+
  geom_vline(aes(xintercept=0), linetype=2, color="grey45")+
  geom_joy(stat="density", adjust=3, alpha=0.8, aes(x=estimate, y=Treatment, fill=Treatment, color=Treatment, height = ..density..))+
  geom_text(data = slope_probs, aes(x=xpos, y=ypos, label = prob_text), size = 3)+
  facet_wrap(~Type, scales = "free")+
  scale_y_discrete(labels = c("Drought","Irrigation"))+
  scale_fill_manual(values = treat_cols)+
  scale_color_manual(values = treat_cols)+
  guides(fill=FALSE, color=FALSE)+
  xlab("Coefficient Value")+
  ylab(NULL)+
  theme_bw()+
  theme(panel.grid.minor = element_blank(),
        panel.border     = element_blank(), 
        axis.line        = element_blank(),
        axis.ticks       = element_blank(),
        strip.background = element_blank(),
        strip.text       = element_text(face="bold"),
        panel.spacing    = unit(1, units = "cm"))
```

\newpage{}

\subsection{Stan Code}
```{r stan1, echo=T, eval=F}
data {
  int<lower=0> Npreds;         # number of covariates, including intercept
  int<lower=0> Npreds2;        # number of random effect covariates
  int<lower=0> Nplots;         # number of plots
  int<lower=0> Ntreats;        # number of treatments
  int<lower=0> Nobs;           # number of observations
  int<lower=0> Nyears;         # number of years
  vector[Nobs] y;              # vector of observations
  row_vector[Npreds] x[Nobs];	 # design matrix for fixed effects
  row_vector[Npreds2] z[Nobs]; # simple design matrix for random effects 
  int plot_id[Nobs];           # vector of plot ids
  int treat_id[Nobs];          # vector of treatment ids
  int year_id[Nobs];           # vector of year ids
}

parameters {
	vector[Npreds] beta;	             # overall coefficients
	vector[Nyears] year_off;           # vector of year effects
	cholesky_factor_corr[Npreds2] L_u; # cholesky factor of plot random effect corr matrix
  vector[Npreds2] beta_plot[Nplots]; # plot level random effects
	vector<lower=0>[Npreds2] sigma_u;  # plot random effect std. dev.
	real<lower=0> sd_y;                # treatment-level observation std. dev.
	real<lower=0> sigma_year;          # year std. dev. hyperprior
	real<lower=0> phi;                 # dispersion parameter
}

transformed parameters {
  vector[Nobs] A;             # parameter for beta distn
  vector[Nobs] B;             # parameter for beta distn
  vector[Nobs] yhat;               # vector of expected values
  vector[Npreds2] u[Nplots];       # transformed plot random effects
  matrix[Npreds2,Npreds2] Sigma_u; # plot ranef cov matrix

  Sigma_u = diag_pre_multiply(sigma_u, L_u); # cholesky factor for plot-level covariance matrix 
  for(j in 1:Nplots)
    u[j] = Sigma_u * beta_plot[j]; # plot random intercepts and slopes

  # regression model for expected values (one for each plot-year)
  for (i in 1:Nobs)
    yhat[i] = inv_logit(x[i]*beta + z[i]*u[plot_id[i]] + year_off[year_id[i]]);

  A = yhat * phi;
  B = (1.0 - yhat) * phi;
}

model {
  ####  PRIORS
  phi ~ cauchy(0, 5); 
  sigma_year ~ cauchy(0,2.5);
  sd_y ~ cauchy(0,2.5);
  year_off ~ normal(0,sigma_year); # priors on year effects, shared variance
  beta ~ normal(0,5);	             # priors on treatment coefficients
  L_u ~ lkj_corr_cholesky(2.0);      # prior on the cholesky factor which controls the 
                                    # correlation between plot level treatment effects
  sigma_u ~ cauchy(0,2.5);

  for(i in 1:Nplots)
		beta_plot[i] ~ normal(0,1); # plot-level coefficients for intercept and slope

	####  LIKELIHOOD
  y ~ beta(A, B); # observations vary according to beta distribution
}

generated quantities{
  corr_matrix[Npreds2] R = multiply_lower_tri_self_transpose(L_u);
  cov_matrix[Npreds2] V = quad_form_diag(R,sigma_u);
}

```

### 2018-01-18 manuscript/usses_water_ms.Rmd
adds SOILWAT appendix and updates ordering in main text. Closes #13

\new{Following the above procedure, we still lacked soil moisture data for March 2012 (observations did not start unitl April) and for a string of days in 2013 during which the soil moisture probes failed take readings.
Therefore, we used a version of the SOILWAT soil moisture model} [@Sala1992a] \new{that has been specifically designed for semiarid shrublands and grasslands} [@Bradford2014].
\new{The model was parameterized using generic sagebrush steppe parameters and local soil texture, soil bulk density, and precipitation data} [@Kleinhesselink2017b].
\new{We used the locally parameterized SOILWAT model to generate daily soil moisture predictions for the duration of the experiment, but only used SOILWAT predictions where there were gaps in our data (Fig. 2B and Appendix X).}

@@

For each plot and year, we averaged the four readings for each wavelength and then calculated a greenness index based on the same bands used to calculate NDVI using the MODIS (Moderate Resolution Imaging Spectroradiometer) and AVHRR (Advanced Very High Resolution Radiometer) bands for NDVI.
We regressed the greenness index against the dry biomass weight from the ten calibration plots to convert the greenness index to ANPP.
We fit regressions to a MODIS-based index and an AVHRR-based index for each year and retained the regression with the better fit based on $R^2$ values.
We then predicted ANPP using the best regression equation for each year (Appendix 2).
\new{Our analysis, described below, can be done using NDVI directly, but we focus on ANPP to more easily relate our findings to previous work and because our results do not change if we analyze NDVI directly (Appendix 4).}

@@

\noindent{}where $\Sigma$ is the variance-covariance matrix and $j(i)$ reads as "plot $j$ associated with observation $i$". 
The random year effects ($\boldsymbol{\eta}$) are drawn from a normal prior with mean 0 and standard deviation $\sigma_{\text{year}}$, which was drawn from a half-Cauchy distribution. 
\new{We used a vague prior distribution for each $\beta$: $\boldsymbol{\beta} \sim \text{Normal}(0,5)$.}
A full description of our model and associated R [@R2016] code is in Appendix 3.

@@

All R code and data necessary to reproduce our analysis has been archived on Figshare (\emph{link here after acceptance}) and released on GitHub (https://github.com/atredennick/usses_water/releases/v0.1).
We also include annotated Stan code in our model description in Appendix 3.

@@

First, it could be that our precipitation manipulations were not large enough to induce a response.
That is, a 50\% decrease in any given year may not be abnormal given our site's historical range of variability [@Knapp2017].
We cannot definitively rule out this possibility, but we have reason to believe our treatments \emph{should} have been large enough.
Using the methods described by @Lemoine2016, we calculated the percent reduction and increase of \new{water-year} precipitation necessary to reach the 1\% and 99\% extremes of the historical precipitation regime at our site (Fig. A5-1).
\new{The 1\% quantile of water-year precipitation at our site is 78 mm, a 26% reduction from the mean, and the 99\% quantile is 545 mm, a 84\% increase from mean growing season precipitation (Appendix 5).
Thus, our drought treatment represented extreme precipitation amounts, and that is the treatment where we observed a small effect on the slope between soil moisture and ANPP (Fig. 3A).
The irrigation treatment may not have been extreme enough to induce a response.}

@@

\begin{description}
\item [Appendix 1.] \new{Additional information on plant community composition and dynamics, Table A1-1, Table A1-2, Table A1-3, Fig. A1-1, Fig. A1-2, Fig. A1-3, Fig. A1-4, Fig. A1-5, and Fig. A1-6.}
\item [Appendix 2.] Additional methods and information on estimating aboveground net primary productivity, Table A2-1, and Table A2-2.
\item [Appendix 3.] Details of the hierarchical Bayesian model, Fig. A2-1, Fig. A2-2, and Fig. A2-3.
\item [Appendix 4.] \new{Results from analysis of NDVI without conversion to ANPP, Fig. A3-1.}
\item [Appendix 5.] Details on analysis of precipitation historical range of variability and Fig. A4-1.
\end{description}

>>>>

\new{Following the above procedure, we still lacked soil moisture data for March 2012 (observations did not start unitl April) and for a string of days in 2013 during which the soil moisture probes failed take readings.
To fill in these gaps, we used a version of the SOILWAT soil moisture model} [@Sala1992a] \new{that has been specifically designed for semiarid shrublands and grasslands} [@Bradford2014].
\new{The model was parameterized using generic sagebrush steppe parameters and local soil texture, soil bulk density, and precipitation data} [@Kleinhesselink2017b].
\new{We used the locally parameterized SOILWAT model to generate daily soil moisture predictions for the duration of the experiment, but only used SOILWAT predictions where there were gaps in our data (Fig. 2B and Appendix 2).
SOILWAT predicted daily soil moisture under ambient conditions similar to our control plots.
We applied the same statistical model and procedure described above to estimate soil moisture in drought and irrigation plots based on control plot conditions.}

@@

For each plot and year, we averaged the four readings for each wavelength and then calculated a greenness index based on the same bands used to calculate NDVI using the MODIS (Moderate Resolution Imaging Spectroradiometer) and AVHRR (Advanced Very High Resolution Radiometer) bands for NDVI.
We regressed the greenness index against the dry biomass weight from the ten calibration plots to convert the greenness index to ANPP.
We fit regressions to a MODIS-based index and an AVHRR-based index for each year and retained the regression with the better fit based on $R^2$ values.
We then predicted ANPP using the best regression equation for each year (Appendix 3).
\new{Our analysis, described below, can be done using NDVI directly, but we focus on ANPP to more easily relate our findings to previous work and because our results do not change if we analyze NDVI directly (Appendix 5).}

@@

\noindent{}where $\Sigma$ is the variance-covariance matrix and $j(i)$ reads as "plot $j$ associated with observation $i$". 
The random year effects ($\boldsymbol{\eta}$) are drawn from a normal prior with mean 0 and standard deviation $\sigma_{\text{year}}$, which was drawn from a half-Cauchy distribution. 
\new{We used a vague prior distribution for each $\beta$: $\boldsymbol{\beta} \sim \text{Normal}(0,5)$.}
A full description of our model and associated R [@R2016] code is in Appendix 4.

@@

All R code and data necessary to reproduce our analysis has been archived on Figshare (\emph{link here after acceptance}) and released on GitHub (https://github.com/atredennick/usses_water/releases/v0.1).
We also include annotated Stan code in our model description in Appendix 4.

@@

First, it could be that our precipitation manipulations were not large enough to induce a response.
That is, a 50\% decrease in any given year may not be abnormal given our site's historical range of variability [@Knapp2017].
We cannot definitively rule out this possibility, but we have reason to believe our treatments \emph{should} have been large enough.
Using the methods described by @Lemoine2016, we calculated the percent reduction and increase of \new{water-year} precipitation necessary to reach the 1\% and 99\% extremes of the historical precipitation regime at our site (Fig. A5-1).
\new{The 1\% quantile of water-year precipitation at our site is 78 mm, a 26% reduction from the mean, and the 99\% quantile is 545 mm, a 84\% increase from mean growing season precipitation (Appendix 6).
Thus, our drought treatment represented extreme precipitation amounts, and that is the treatment where we observed a small effect on the slope between soil moisture and ANPP (Fig. 3A).
The irrigation treatment may not have been extreme enough to induce a response.}

@@

\begin{description}
\item [Appendix 1.] \new{Additional information on plant community composition and dynamics, Table A1-1, Table A1-2, Table A1-3, Fig. A1-1, Fig. A1-2, Fig. A1-3, Fig. A1-4, Fig. A1-5, and Fig. A1-6.}
\item [Appendix 2.] \new{Additional details on soil moisture modeling with SOILWAT and Fig. A2-1.}
\item [Appendix 3.] Additional methods and information on estimating aboveground net primary productivity, Table A3-1, and Table A3-2.
\item [Appendix 4.] Details of the hierarchical Bayesian model, Fig. A4-1, Fig. A4-2, and Fig. A4-3.
\item [Appendix 5.] \new{Results from analysis of NDVI without conversion to ANPP, Fig. A5-1.}
\item [Appendix 6.] Details on analysis of precipitation historical range of variability and Fig. A6-1.
\end{description}
